Grade Retention and Mixed Race

### 2022-09-09 products/paper/main.qmd
Edit explanation of example figure for new design

@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly behind expected grade. My goal is to understand the degree of grade retention for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement.

As an example, I display the percent of students clearly behind grade for White, Black, and Black/White students in @fig-example. The point estimate and confidence interval for the Black/White group are shown in red, while the two monoracial groups are shown via grey dots and confidence bands. For all three cases, I use 83.4% confidence intervals, because overlap on such intervals indicates that the observed difference between the two measures in the sample is not statistically distinguishable from zero in the population at $p<.05$ [@knol_mis_2011].[^2] In this case, the lack of overlap between any of the confidence bands indicates that the Black/White probability of being clearly behind expected grade is statistically distinguishable from both monoracial probabilities.

[^2]: The correspondence to $p<.05$ is only approximate due to differences in the standard errors between groups. Calculating more precise intervals requires that those intervals vary across each pairwise comparison which does not work for the visualizations here. The approximation will be very close in all cases used here.

```{r example-plot}
#| label: fig-example
#| fig-cap: Probability of being clearly below expected grade for biracial black/white respondents in comparison to their monoracial comparison groups. 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level.

@@

Additionally, I include a measure of the "halfway" point between the two racial groups.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups. In this case, Black/White students are much closer to White students in their risk of grade retention, and have probabilities much lower than both Black students and the halfway expectation.

[^3]: The halfway point is given by taking the mean between the two monoracial group. Its standard error is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups.

@@

The Black/White case stands out as an outlier. Black/White students have similar resources as Black students and both groups have fewer resources than White students. This is observable in @tbl-resources across family income, home ownership, and parental educational attainment. Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions. 
The effect of controlling for these resources will be complex and different across biracial groups. For those biracial groups with resources similar to their more advantaged constituent group, we should see that these controlling for these resources should reduce their advantage relative to the less advantaged monoracial group. For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites. For the remaining “in-between” groups, it is more difficult to make a priori predictions.

>>>>

@tbl-descriptive shows the percent of each monoracial and biracial group that is clearly behind expected grade. My goal is to understand the risk of being behind expected grade for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement.

As an example, I display the percent of students clearly behind expected grade for White, Black, and Black/White students in @fig-example. I display confidence bands/bars around all three point estimates, but these confidence ranges require some explanation as they are not standard 95% confidence intervals. My goal is to determine whether the biracial group's point estimate is statistically distinguishable from a given monoracial constituent group's point estimate by non-overlap of confidence intervals. Contrary to common belief, overlap in 95% confidence intervals is not equivalent to failing a hypothesis test of difference at $p<0.05$ and leads to far more stringent tests [@knol_mis_2011]. When standard errors between groups are equal, overlap in 83.4% confidence intervals will achieve this goal. However, in cases where standard errors vary substantially across estimates, the calculation of the appropriate confidence interval is more complex and varies by each pairwise comparison.[^2]

[^2]: The formula for calculating the z-score for the correct interval is given by $$1.96 * \frac{\sqrt{1+\rho^2}}{1+\rho}$$ where $\rho$ is the ratio of the standard errors for the two statistics. See @knol_mis_2011 for detailed derivation of this formula.

```{r example-plot}
#| label: fig-example
#| fig-cap: Probability of being clearly behind expected grade for biracial black/white respondents in comparison to their monoracial comparison groups. Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level.

@@

I calculate two confidence intervals for each biracial group. Each confidence interval is in comparison to one of the constituent monoracial groups. These confidence intervals are color-coded in @fig-example to indicate the reference group. I also only draw half-intervals in the direction of the monoracial group's point estimate. Overlap in these color corresponding bars and bands indicates that the difference between the biracial group and the monoracial group is not statistically significant at $p<.05$. For example, the yellow bar shown in @fig_example for biracial Black/White students does not quite overlap with the yellow band for White students, indicating that these point estimates for these two groups are statistically distinguishable at $p<.05$. <!-- TODO provide actual p-value for comparison -->

Additionally, I include a measure of the "halfway" point between the two monoracial groups in dark grey.[^3] This halfway point allows me to determine where the biracial group falls relative to the expectation of being halfway between the two constituent monoracial groups. In this case, Black/White students are much closer to White students in their risk of grade retention, and have probabilities much lower than both Black students and the halfway expectation.

[^3]: The halfway point is given by taking the mean between the two monoracial group. Its standard error is given by $\sqrt{(s_1^2+s_2^2)/2}$ where $s_1$ and $s_2$ are the standard errors of the estimates for the two monoracial groups. To aboid clutter, I do not draw a third half-confidence interval on the biracial estimate itself, but this estimate will always be roughly about halfway between the length of the other two intervals.

@@

The Black/White case stands out as an outlier. Black/White students have similar resources as Black students and both groups have fewer resources than White students. This is observable in @tbl-resources across family income, home ownership, and parental educational attainment. Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions. The effect of controlling for these resources will be complex and different across biracial groups. For those biracial groups with resources similar to their more advantaged constituent group, we should see that these controlling for these resources should reduce their advantage relative to the less advantaged monoracial group. For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites. For the remaining "in-between" groups, it is more difficult to make a priori predictions.
